Geometric variational problems

### 2020-11-11 NSF2020/Description/specificaim1_ver2.tex
rjr
yny 1111

%\begin{figure}
%\begin{center}
%\includegraphics[width=0.9\textwidth]{figures/SA1_fig1.pdf}
%\end{center}
%\caption{\footnotesize (A--C) The splay ($\Div \mathbf{d}$), 
%saddle splay ($\det \mathsf{D}$) and twist ($\Curl \mathbf{d}$) elastic distortions of 
%a monolayer. (D) The monolayer neutral surface $\Sigma$,  
%director $\mathbf{d}$ and unit normal $\mathbf{n}$ in local coordinates.
%(E--F) Lipids are able to tilt from  the surface normal, and stretch.
%(H--J) Intermediates of membrane fusion and (K) experimental image of a stalk \cite{Aeffner2012}. }
%\label{fig:distortions}
%\end{figure}
The goal of Specific Aim 1 is characterize the material properties of many-body, self-assembled amphiphiles.
For amphiphiles assembled into bilayers, these properties are described by membrane continuum mechanics.
Our goal is to be able to map the parameters of the particle-based model onto the elastic moduli from continuum theory.
Results from this goal will facilitate simulators to use the hydrophobic attraction force calculations
to model bilayers with specific composition. These calculations have provably less computational complexity than
those of molecular dynamics simulations and possess the molecular granularity lacking from continuum models.

The modern theory of membrane continuum mechanics 
was pioneered by Hamm and Kozlov (HK) \cite{Hamm2000}.
The theory is widely used to describe biological phenomona 
fission \cite{FrEsAkSh15, Maetal15, PhysRevE.79.031926},
fusion \cite{ChKo08, KoKo2002,Kuzmin7235,Aeffner2012},
poration \cite{Gaetal20}, phase boundaries and interaction with inclusions
\cite{SeLeMaEg17,Saetal20, Pietal20}, which require resolution of the internal structure of the membrane.  
Using the HK  model as a base,
PI Ryham and collaborators calculated, for the first time in continuum theory, a least energy path for transitions between planar bilayers, a membrane stalk,
hemifusion diaphragm and the fusion pore \cite{RyWaCo13,RyKlYaCo16},
and the energies calculated by these continuum studies (Figure \ref{fig:barriers}A),
are in agreement with barrier heights derived by molecular dynamics and experimental studies \cite{FrRoPi17}.

Recently, there has been a revival in interest in the HK theory. The quadratic assumption
for the elasticity energy density has caused researchers to question the applicability of \eqref{ansatz3} for large curvatures \cite{PhysRevLett.117.188102, ARGUDO20161619}. 
The article \cite{C9SM02079A} counters that energies derived from molecular dynamics and elasticity agree, even when curvatures are large.
This proposal will bring mathematical analysis to some of the assumptions 
that went into the HK theory.

\begin{figure}
  \centering{\includegraphics[width=0.80\textwidth]{Figures/Deformations.pdf}}
  \caption{\label{fig:deformations}
    }
\end{figure}

The HK framework assumes a three-dimensional, internal structure for lipid monolayers.
The internal structure consists of straight fibers that represent elongated hydrocarbon chains, and these fibers
tilt and stretch with respect to the dividing surface--the surface formed lying between the hydrocarbon chains and polar heads of lipids.
Accordingly, the deformation takes the form 
\begin{equation}
  \label{LMdeformation}
{\bf x}(X_1, X_2, X_3) = {\bf x}_0(X_1, X_2) + \zeta(x_0(X_1, X_2), X_3) {\bf n}(x_0(X_1, X_2))
\end{equation}
where $(X_1,X_2,X_3)$ are points in a reference volume.
The map $x_0$ parametrizes the dividing surface $\Sigma$, and the unit vector field ${\bf n}$ parametrizes the
direction of the lipid tails in the deformed state \cite{doi:10.1021/jp075641w,KLAUDA20083074}.
The parameter $X_3$ and the function $\zeta$ parametrizes the distance along the hydrocarbon chain in the reference state
and the deformed state, respectively, and so $0 \leq X_3 \leq \delta$ where $\delta = $ 2.5 nm is a realistic monolayer thickness.

The elastic theory assumes an energy density that is quadratic in the Green-Lagrange strain tensor for ${\bf x}$. 
The chains stretch and compress to satisfy incompressibility.
Symmetries with respect to mirror reflection and in-plane rotation, the internal structure assumptions and incompressibility
lead to an elastic surface energy over the dividing surface $\Sigma$. 
This energy decomposes into four, fundamental and independent deformations:
splay ($\Div {\bf n}$), twist ($\Curl {\bf n}$), saddle splay ($\det \nabla {\bf n}$) and tilt ${\bf t}$:
\begin{equation}
\label{ansatz3}
\begin{aligned}
\int_{\Sigma} 
  \tfrac{1}{2}\KB\left[ \left( \Div {\bf n} + k_0\right)^2 - k_0^2\right] 
+ \tfrac{1}{2}\KT (\Curl {\bf n})^2 + \KG  \det \nabla {\bf n} + \tfrac{1}{2}\KTH |{\bf t}|^2 \,dA.
\end{aligned}
\end{equation}
Here, $(\nabla {\bf n})_{ij} = \partial_{x_j}{\bf n} \cdot {\bf e}_i$ is the surface gradient with
orthonormal tangent vectors $\{{\bf e}_1, {\bf e}_2\}$.
\todo[inline]{YNY and BQ: What do you call a coordinate system where the metric is the identity at one point?}
Also, the surface divergence 
$\Div {\bf n}$ is the trace, and the surface curl $\Curl {\bf n}$ is the anti-trace (sum of the off-diagonal elements), of $\nabla {\bf n}$ respectively.
respectively (Figure \ref{fig:distortions}A--C).
The deformations come with elastic coefficients: the bending modulus $\KB$, twist modulus $\KT$ 
and saddle-splay modulus $\KG$ and tilt modulus $\KTH$
HK define the tilt vector ${\bf t} = {\bf n}/({\bf N}\cdot {\bf n}) - {\bf N}$ where ${\bf N}$ is the unit surface normal.

The parameter $k_0$ is the  spontaneous curvature and it determines the preferred lipid splay \cite{RoLi15,Kozlov2007}. 
As an illustration, the lipid DSPC has $k_0 = -0.1$ nm$^{-1}$ 
%as a result of  %having a relatively long acyl chain. 
and these lipids would line the  inner monolayer of a 
spherical liposome, because the addition of a positive splay $\Div \mathbf{n}$  in \eqref{ansatz3}
to the negative spontaneous curvature $k_0$ leads to lower energy \cite{Kamal22245, C3SM51829A, RoLi15,FriedSeguin15}.
The subtraction of $k_0^2$ makes the energy distortion free
\cite{Helfrich73,PhysRevLett.113.248102,Hamm2000}.

Under spatial scales much larger than the
membrane thickness, membrane energy is well-characterized by the
Canham-Helfrich energy used throughout the fluid-structure literature
\cite{QiangDu09, Lowengrub07,KimLai2010_JCP, Hu, HuLaiSeolEtAl2016_JCP,
  qua-bir2014, qua-vee-you2019}.
\todo[inline]{YNY: place your recent works here}
The Canham-Helfrich energy is actually a special case of
\eqref{ansatz3} obtained by setting ${\bf n} =  \pm {\bf N}$ (the $\pm$ depending on
orientation) and collapsing both monolayers onto the membrane midplane.
There, splay $\Div n$ equal to twice the mean curvature, saddle splay
$\det \nabla n$ equal to the Gauss curvature, and both twist and tilt
are equal to zero.

@@

%\begin{center}
%Anticipated model simulations
%\end{center}
\subsubsection{Anticipated model simulations}
To characterize material properties, we compare the energies and forces generated by
a collection of amphiphilic particles against the energies and forces of a continuum bilayer.
The specific goals are {\bf (i)} to assess whether the energies \eqref{ansatz3} are the same
for all reasonable deformation and if this is the case {\bf (ii)} to determine the value of the elastic $\KB$,
$\KT$, $\KG$ and $\KTH$; {\bf (iii)} us to understand how the model parameters $\rho,$ $\gamma$, and particle shape map onto elastic moduli.

Motivated by our preliminary results \cite{Fu2018_SIAM}
we use realistic values for particle diameter $2$ nm \cite{Boal},
screening length $\rho = 2.5$ nm \cite{Eriksson1989,Lin2005,Parsegian,Israelachvili80,TerziDeserno17}
and interfacial tension $\gamma=4.1$ pN nm$^{-1}$ \cite{GarciaSaez, KUZMIN2005, Petelska2012, Jackson}
model parameters. The repulsion strength, repulsion length scale, particle shape (e.g. disks, ellipses), 
particle diameter and the hydrophobic boundary condition \eqref{SL}
are additional parameters defining the coarse-grained representation.

>>>>

%\begin{figure}
%\begin{center}
%\includegraphics[width=0.9\textwidth]{figures/SA1_fig1.pdf}
%\end{center}
%\caption{\footnotesize (A--C) The splay ($\Div \mathbf{d}$), 
%saddle splay ($\det \mathsf{D}$) and twist ($\Curl \mathbf{d}$) elastic distortions of 
%a monolayer. (D) The monolayer neutral surface $\Sigma$,  
%director $\mathbf{d}$ and unit normal $\mathbf{n}$ in local coordinates.
%(E--F) Lipids are able to tilt from  the surface normal, and stretch.
%(H--J) Intermediates of membrane fusion and (K) experimental image of a stalk \cite{Aeffner2012}. }
%\label{fig:distortions}
%\end{figure}
The goal of Specific Aim 1 is to characterize the material properties of many-body, self-assembled amphiphiles.
For amphiphiles assembled into bilayers, these properties are described by membrane continuum mechanics.
Our goal is to be able to map the parameters of the particle-based model onto the elastic moduli from continuum theory.
Results from this goal will facilitate simulators to use the hydrophobic attraction force calculations
to model bilayers with specific composition. These calculations have provably less computational complexity than
those of molecular dynamics simulations and possess the molecular granularity lacking from continuum models.

Hamm and Kozlov (HK) pioneered 
the modern theory of membrane continuum mechanics \cite{Hamm2000}, and their theory
%was pioneered by Hamm and Kozlov (HK) \cite{Hamm2000}. The theory 
is widely used to describe biological phenomona 
fission \cite{FrEsAkSh15, Maetal15, PhysRevE.79.031926},
fusion \cite{ChKo08, KoKo2002,Kuzmin7235,Aeffner2012},
poration \cite{Gaetal20}, phase boundaries and interaction with inclusions
\cite{SeLeMaEg17,Saetal20, Pietal20}, which require resolution of the internal structure of the membrane.  
%Using the HK  model as a base,
%PI RR and collaborators calculated, for the first time in continuum theory, a least energy path for transitions between planar bilayers, a membrane stalk,
%hemifusion diaphragm and the fusion pore \cite{RyWaCo13,RyKlYaCo16},
%and the energies calculated by these continuum studies (Figure \ref{fig:barriers}A),
%are in agreement with barrier heights derived by molecular dynamics and experimental studies \cite{FrRoPi17}.  
%
Recently, there has been a revival in interest in the HK theory as the quadratic assumption
for the elasticity energy density has caused researchers to question the applicability of \eqref{ansatz3} for large curvatures \cite{PhysRevLett.117.188102, ARGUDO20161619}. 
For example, Galimzyanov {\it et al.} \cite{C9SM02079A} countered that energies derived from molecular dynamics and elasticity agree, even when curvatures are large.
This proposal will develop much-needed mathematical analysis to resolve these controversies due to the assumptions in the HK theory.

\begin{figure}
  \centering{\includegraphics[width=0.80\textwidth]{Figures/Deformations.pdf}}
  \caption{\label{fig:deformations}Sketch of the HK membrane model \cite{Hamm2000}.
    }
\end{figure}

The HK framework assumes a three-dimensional lipid monolayers with internal structure
consisting of straight fibers that represent elongated hydrocarbon chains. HK assumed an
elastic energy density ${\cal W}$ quadratic in the Green-Lagrange strain tensor for membrane deformation ${\bf x}$.
%
%
%
%These fibers
%tilt and stretch with respect to the dividing surface--the surface formed lying between the hydrocarbon chains and polar heads of lipids.
%Accordingly, the deformation takes the form 
%\begin{equation}
%  \label{LMdeformation}
%{\bf x}(X_1, X_2, X_3) = {\bf x}_0(X_1, X_2) + \zeta(x_0(X_1, X_2), X_3) {\bf n}(x_0(X_1, X_2))
%\end{equation}
%where $(X_1,X_2,X_3)$ are points in a reference volume.
%The map $x_0$ parametrizes the dividing surface $\Sigma$, and the unit vector field ${\bf n}$ parametrizes the
%direction of the lipid tails in the deformed state \cite{doi:10.1021/jp075641w,KLAUDA20083074}.
%The parameter $X_3$ and the function $\zeta$ parametrizes the distance along the hydrocarbon chain in the reference state
%and the deformed state, respectively, and so $0 \leq X_3 \leq \delta$ where $\delta = $ 2.5 nm is a realistic monolayer thickness. 
%
%The elastic theory assumes an energy density that is quadratic in the Green-Lagrange strain tensor for ${\bf x}$. 
%The chains stretch and compress to satisfy incompressibility.
%Symmetries with respect to mirror reflection and in-plane rotation, the internal structure assumptions and incompressibility
%lead to an elastic surface energy over the dividing surface $\Sigma$. 
%
This energy decomposes into four, fundamental and independent deformations  (figure~\ref{fig:deformations}):
splay ($\Div {\bf n}$), twist ($\Curl {\bf n}$), saddle splay ($\det \nabla {\bf n}$) and tilt ${\bf t}={\bf n}/({\bf N}\cdot {\bf n}) - {\bf N}$ where ${\bf N}$ is the unit surface normal:
\begin{equation}
\label{ansatz3}
\begin{aligned}
{\cal W} \equiv \int_{\Sigma} 
  \tfrac{1}{2}\KB\left[ \left( \Div {\bf n} + k_0\right)^2 - k_0^2\right] 
+ \tfrac{1}{2}\KT (\Curl {\bf n})^2 + \KG  \det \nabla {\bf n} + \tfrac{1}{2}\KTH |{\bf t}|^2 \,dA.
\end{aligned}
\end{equation}
Here, 
%$(\nabla {\bf n})_{ij} = \partial_{x_j}{\bf n} \cdot {\bf e}_i$ is the surface gradient with
%orthonormal tangent vectors $\{{\bf e}_1, {\bf e}_2\}$.
%%\todo[inline]{YNY and BQ: What do you call a coordinate system where the metric is the identity at one point?}
%Also, the surface divergence 
%$\Div {\bf n}$ is the trace, and the surface curl $\Curl {\bf n}$ is the anti-trace (sum of the off-diagonal elements), of $\nabla {\bf n}$ respectively.
%respectively (Figure \ref{fig:distortions}A--C).
the deformations come with elastic coefficients: the bending modulus $\KB$, twist modulus $\KT$ 
and saddle-splay modulus $\KG$ and tilt modulus $\KTH$
%HK define the tilt vector ${\bf t} = {\bf n}/({\bf N}\cdot {\bf n}) - {\bf N}$ where ${\bf N}$ is the unit surface normal.
%
The parameter $k_0$ is the  spontaneous curvature and it determines the preferred lipid splay \cite{RoLi15,Kozlov2007}. 
As an illustration, the lipid DSPC has $k_0 = -0.1$ nm$^{-1}$ 
%as a result of  %having a relatively long acyl chain. 
and these lipids would line the  inner monolayer of a 
spherical liposome, because the addition of a positive splay $\Div \mathbf{n}$  in \eqref{ansatz3}
to the negative spontaneous curvature $k_0$ leads to lower energy \cite{Kamal22245, C3SM51829A, RoLi15,FriedSeguin15}.
The subtraction of $k_0^2$ makes the energy distortion free
\cite{Helfrich73,PhysRevLett.113.248102,Hamm2000}.

Under spatial scales much larger than the
membrane thickness, membrane energy is well-characterized by the
Canham-Helfrich energy used throughout the fluid-structure literature
\cite{QiangDu09, Lowengrub07,KimLai2010_JCP, Hu, HuLaiSeolEtAl2016_JCP,
  qua-bir2014, qua-vee-you2019}.
%\todo[inline]{YNY: place your recent works here}
The Canham-Helfrich energy is actually a special case of
\eqref{ansatz3} obtained by setting ${\bf n} =  \pm {\bf N}$ (the $\pm$ depending on
orientation) and collapsing both monolayers onto the membrane midplane.
%There, splay $\Div n$ equal to twice the mean curvature, saddle splay
%$\det \nabla n$ equal to the Gauss curvature, and both twist and tilt
%are equal to zero.

@@

%\begin{center}
%Anticipated model simulations
%\end{center}
\subsubsection{Anticipated model simulations}
To characterize material properties of a bilayer, we compare the energies and forces generated by
a collection of amphiphilic particles against the energies and forces of a continuum bilayer.
We propose to first {\bf (i)} Assess whether the continuum energies \eqref{ansatz3} are the same
as the corresponding energies in the particle system for all reasonable deformation.
If this is the case we will {\bf (ii)} determine the value of the elastic $\KB$,
$\KT$, $\KG$ and $\KTH$, and then {\bf (iii)} understand how the model parameters $\rho,$ $\gamma$, and particle shape map onto elastic moduli. 
%
%Motivated by our preliminary results \cite{Fu2018_SIAM}
%we use realistic values for particle diameter $2$ nm \cite{Boal},
%screening length $\rho = 2.5$ nm \cite{Eriksson1989,Lin2005,Parsegian,Israelachvili80,TerziDeserno17}
%and interfacial tension $\gamma=4.1$ pN nm$^{-1}$ \cite{GarciaSaez, KUZMIN2005, Petelska2012, Jackson}
%model parameters. The repulsion strength, repulsion length scale, particle shape (e.g. disks, ellipses), 
%particle diameter and the hydrophobic boundary condition \eqref{SL}
%are additional parameters defining the coarse-grained representation.
